A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-10-16 model.tex
edit tex

In this section we review the basic properties of the 
single-fluid mixture using Eqs.\ref{masseq}--\ref{momeq} with 
Eq. \ref{eff_energy} in place of the dust continuity equation. Our
discussion is based on the corresponding axisymmetric linearized
equations. A variable $f$ is
subject to Eulerian perturbations of the form 
\begin{align}
  \delta f(r,z)\exp{\left(-\ii\sigma t\right)},
\end{align}
where $\sigma$ is the complex mode frequency. In Appendix
\ref{var_prin} we derive the integral relation,  
\begin{align}
&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV.\label{int_rel}
\end{align} 
The real coefficients $A,B,C,D$ can be read off 
Eq. \ref{integral_ex}. Note that $B=C$ from dynamical equilibrium
(Eq. \ref{vshear}). We now consider various limits of
Eq. \ref{int_rel}.

@@

%except here the entropy $s$ is really a
%representation of the dust distribution via $s=\ln{\left[c_s^2\left(1
%    - \tepsilon\right)\right]}$.

%These criteria are applicable to any adiabatic fluid. 
%For an isothermal mixture, we may take $s=\ln{(1-\tepsilon)}$.   
%from the definition of $s$ and  
%the equation of state (Eq. \ref{eos}) with uniform sound-speed. 
%\begin{align}
%  s \equiv \ln \frac{P}{\rho} = \ln{(1-\tepsilon)} 
%\end{align}
% The first equality is
%similar to the usual defintion for gas entropy (with unit adiabatic
%index). The second follows from the imposed equation of state,
%Eq. \ref{eos}, with a constant sound-speed. 
%The stability conditions are thus
%In accretion disks, $\rhog$, $\rhod$ and $\tepsilon$ are expected to
%decrease away from the mid-plane
%\begin{align*}  
%  \frac{\p\rhog}{\p |z|} < 0,
%, \quad \frac{\p \rhog}{\p r} < 0,   
%\end{align*}
%and similarly for $\rhod$, $\tepsilon$, and the dust-to-gas ratio $\epsilon$. 
%We may thus evaluate the stability criteria at $z>0$ without loss of
%generality.

%We insert the vertical shear profile,
%Eq. \ref{vshear}, into Eq. \ref{dusty_solberg} and write the result 
%in terms of $\epsilon$, to find

@@

Eq. \ref{stability_est1}---\ref{stability_est2} are easily satisfied
in thin accretion disks where radial gradients are $O(1/r)$, 
$h_\mathrm{g},\epsilon,\ll 1$ and $\Htilde/H_\mathrm{g}$ not too
large.  
Then the
magnitude of the second term on the left-hand-side  of either
inequality is much less than unity. Thus radially smooth, strictly
isothermal disks perfectly coupled with dust are stable against
axisymmetric perturbations.

@@

\subsection{Thermodynamics of dust-drag instabilities}\label{dust_work}
Now consider constant $c_s$ but $\tstop\neq0$. 
%When $c_s(r,z)$ is non-uniform but $\tstop=0$, 
Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, so
we may have growing oscillations. Writing $\sigma = -\omega + \ii s$
where $\omega$ and $s$ are the real frequency and growth rates, 
respectively, the imaginary part of Eq. \ref{int_rel} gives 
\begin{align}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
\end{align}
%In this case 
%the effective energy equation is 
%\begin{align}\label{dusty_cooling}
%  \frac{DP}{Dt} &= - P \nabla \cdot \bm{v} + \mathcal{C}
%\end{align} 
%The function $\mathcal{C}$ can be considered as a cooling term. 
%By
%analogy with enabling VSI in an adiabatic gas through cooling
%\citep{nelson13,lin15}, we may expect a `dusty VSI' to develop 
%if this term is large enough (physically meaning diffusion of the
%dust-fraction is strong). 
%Note that $\mathcal{C}$ is also responsible for the streaming
%instability  in the strong-drag limit \citep{laibe14}, though this
%does not require vertical stratification. In the form of
%Eq. \ref{dusty_cooling}, however, the streaming instability has a
%thermodynamical interpretation, similar to thermal instabilities in
%stars \citep{latter06}.  
which represent instabilities due to dust-gas friction. 
%Our formalism
%gives these dusty instabilities a thermodynamical interpretation. 
% We further discuss this concept in \S\ref{dust_as_thermo}. 
%In \S\ref{dust_work} we showed that instabilities may 
%arise from the term proportional $\imag\left( \nabla\cdot\dd\bm{v}^*
%\delta C\right)$. 
 Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
may be thought of as a correlation between compression/expansion and  
heating/cooling.

It is well-known that such correlations can lead to pulsational
instabilities in stars \citep{cox67}. One may thus interpret 
dust-drag instabilities in a similar way. We give a
qualitative 
description below, based on the more rigorous treatments of stellar 
pulsations by 
\cite{cox67} and lecture notes from \cite{samadi15} and
J. Christensen-Dalsgaard\footnote{\url{http://astro.phys.au.dk/$\sim$jcd/oscilnotes/}}.

@@

The Lagragian perturbation $\Delta$ of a variable $f$ is 
related to its Eulerian perturbation by $\Delta f = \delta f + 
\bm{\xi}\cdot\nabla f$; where $\bm{\xi}$ is the Lagragian  
displacement ($    \xi_{x,z} =  \ii \dd v_{x,z}/\sigma$). 
%Note that $\Delta
%\mathcal{C} = \delta \mathcal{C}$ for a uniform background $C$ as
%required for equilibrium. 
Then Eq. \ref{lin_mass_full}---\ref{lin_energy_full} imply  
\begin{align} 
  \imag \left(\delta \mathcal{C}
  \nabla\cdot\dd \bm{v}^*\right) = 
  -|\sigma|^2\imag \left(\Delta P 
  \frac{\Delta\rho^*}{\rho}\right). 
\end{align}
This reflects `$TdS  =  PdV$' from thermodynamics, if the 
internal energy is constant. That is, dust-gas drag provides 
heating ($TdS$) to do $PdV$ work. Whether positive or negative work is 
done depends on the correlation between the pressure and density
perturbations, as follows.

>>>>

In this section we review the basic properties of the 
single-fluid mixture using Eqs.\ref{masseq}--\ref{momeq} with 
Eq. \ref{eff_energy} in place of the dust continuity equation. Our
discussion is based on the corresponding axisymmetric linearized
equations. A variable $f$ is
subject to Eulerian perturbations of the form 
\begin{align}
 \real\left[ \delta f(r,z)\exp{\left(-\ii\sigma t\right)},\right]
\end{align}
where $\sigma$ is the complex mode frequency. In Appendix
\ref{var_prin} we derive the integral relation,  
\begin{align}
&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV.\label{int_rel}
\end{align} 
The real coefficients $A,B,C,D$ can be read off 
Eq. \ref{integral_ex}. Note that $B=C$ from dynamical equilibrium
(Eq. \ref{vshear}). We now consider various limits of
Eq. \ref{int_rel}.

@@

@@

Eq. \ref{stability_est1}---\ref{stability_est2} are easily satisfied
in thin accretion disks where radial gradients are $O(1/r)$, 
$h_\mathrm{g}\ll 1$ and $\Htilde/H_\mathrm{g}$ not too
large.  Then the
magnitude of the second term on the left-hand-side  of either
inequality is much less than unity. Thus radially smooth, strictly
isothermal disks perfectly coupled with dust are stable against
axisymmetric perturbations.

@@

\subsection{Thermodynamics of dust-drag instabilities}\label{dust_work}
Now consider constant $c_s$ but $\tstop\neq0$. 
%When $c_s(r,z)$ is non-uniform but $\tstop=0$, 
Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, so
we may have growing oscillations. Writing $\sigma = -\omega + \ii s$
where $\omega$ and $s$ are the real frequency and growth rates, 
respectively, the imaginary part of Eq. \ref{int_rel} gives 
\begin{align}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
\end{align}
which represent instabilities due to dust-gas friction. 
Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
may be thought of as a correlation between compression/expansion and  
heating/cooling.

It is well-known that such correlations can lead to pulsational
instabilities in stars \citep{cox67}. One may thus interpret 
dust-drag instabilities in a similar way. We give a
qualitative 
description below, adapted from the more rigorous treatments of stellar 
pulsations by 
\cite{cox67} and lecture notes by \cite{samadi15} and
J. Christensen-Dalsgaard\footnote{\url{http://astro.phys.au.dk/$\sim$jcd/oscilnotes/}}.

@@

\subsubsection{Work done by dusty gas}

The Lagragian perturbation $\Delta$ of a variable $f$ is 
related to its Eulerian perturbation by $\Delta f = \delta f + 
\bm{\xi}\cdot\nabla f$; where $\bm{\xi}$ is the Lagragian  
displacement ($    \xi_{x,z} =  \ii \dd v_{x,z}/\sigma$). 
%Note that $\Delta
%\mathcal{C} = \delta \mathcal{C}$ for a uniform background $C$ as
%required for equilibrium. 
Then Eq. \ref{lin_mass_full}---\ref{lin_energy_full} imply  
\begin{align} 
  \imag \left(\delta \mathcal{C}
  \nabla\cdot\dd \bm{v}^*\right) = 
  -|\sigma|^2\imag \left(\Delta P 
  \frac{\Delta\rho^*}{\rho}\right), \label{pdv}
\end{align}
which states that `PdV' work (right hand side) is
provided by dust-drag friction. This can be seen explicitly as
follows.

The average rate of work done by the dusty-gas is
\begin{align}
  \mathcal{W} = \frac{1}{T_p}\int^{t+T_p}_{t}dt^\prime\int_M P
  \frac{Dv}{Dt^\prime} dm 
\end{align}
\citep{cox67}, 
where $T_p$ is the oscillation period, $v=1/\rho$ is the specific
volume of the mixture, and the spatial integral is taken over the
total mass $M$ of the mixture and $dm$ is the mass element.

Our aim is to relate the work done by the mixture, $\mathcal{W}$, 
to the growth rate of oscillation amplitudes, $s$. To do this, we
first evaluate $\mathcal{W}$ assuming pure oscillations (perturbations
with time dependence $e^{\ii\omega t}$ and real $\omega$), then show
that in fact the amplitudes grow if the work done is positive.

Noting that only products in the perturbations contribute to 
$\mathcal{W}$, we find that for real $\omega$: 
\begin{align}
  \mathcal{W} = - \frac{\omega}{2} \int \imag\left(\Delta P
  \frac{\Delta \rho^*}{\rho}\right)dV,\label{work_real}
\end{align}
where the integral is taken over the volume of the
mixture. Eq. \ref{work_real} show that a phase difference between
pressure and density leads to work done.

Now combining Eq. \ref{work_real}, \ref{pdv} and \ref{thermal_instability}
gives  
\begin{align}
s = \frac{2}{\mathcal{I}^2}\mathcal{W}, \label{growth_work}
\end{align}
where $\mathcal{I}^2>0$  is the integral in the denominator of
Eq. \ref{thermal_instability}, and we have set $\sigma= - \omega$ in
Eq. \ref{pdv} for pure oscillations.  Thus growing modes in fact arise if
positive work is done by the mixture during an oscillation. This in
turn requires pressure to lag density, discussed next.

\subsubsection{Physical picture of dust-drag instabilities}
